---
title: "71-phams"
title-block-banner: true
author:
  - name: Jinlong Ru
    orcid: 0000-0002-6757-6018
toc: true
format: gfm
bibliography: refs.bib
params:
  name: "71-phams"
---

Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.

```{r setup, message=FALSE, include=FALSE}
here::i_am(paste0(params$name, ".qmd"), uuid = "7edde54d-48ab-4b97-a09d-67e426ba91b0")
projthis::proj_create_dir_target(params$name, clean = FALSE)
path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
path_data <- path_source("00-raw", params$name)
dir.create(path_data, recursive = T)
```

The purpose of this document is ...

```{r packages}
library(here)
```

### Read phams

This phams were created using all datasets, which includes,

1. IMGVR
2. P0062 virome
3. NCBI
4. PHROGs
5. Isolated phages
6. Isolated bacteria
7. ...

::: {.callout-warning}
## Not used in publication

This dataset is not used in the final publication.
:::

```{python}
import pandas as pd
import os
from Bio import SeqIO

def create_orf2family_with_seq(fin_orf2family, fin_faa):
    orf2family = pd.read_csv(fin_orf2family, sep="\t")
    recs = list(SeqIO.parse(fin_faa, "fasta"))
    seqs = []
    for rec in recs:
        seqs.append([rec.id, str(rec.seq)])
    seqs = pd.DataFrame(seqs, columns=["orf", "seq"])
    df_orf2family = pd.merge(orf2family, seqs, on="orf", how="inner")
    return df_orf2family


fin_orf2family = os.path.join(r.path_data, "orf2subfamily2family.tsv")
fin_faa = os.path.join(r.path_data, "all_fam.faa")

df_orf2family = create_orf2family_with_seq(fin_orf2family, fin_faa)
fout = os.path.join(r.path_target(), "orf2family_with_seq.tsv")
df_orf2family.to_csv(fout, sep="\t", index=False)
```


## Tasks

The first task is ...

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
